State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-05-07 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
edited parameter and state estimation section

The dynamic field must be spatially oversampled by using a larger number of sensor then imposed by frequency considerations, to use frequency analysis to determine the minimum distance and width of the basis functions. Following this, the distance between basis functions is usually larger then the distance between sensors, although this is not a requirement for using this estimation framework. For a spatially homogeneous isotropic field the basis functions can be placed on a regular square grid. Following this, the knowledge of the distance between basis functions directly implies the total number of basis functions required.

% 
% Due to the high dimensionality of the brain and the current electrode systems, we can not expect to have more sensors than basis functions.
\section{State and Parameter Estimation}\label{StateAndParameterEstimationSection} In this section we describe  estimating the states that scale the field basis functions, $\mathbf{x}_t$, the connectivity kernel parameters, $\boldsymbol \theta$, and the synaptic time constant, $\xi$. The estimation process consists of a two part iterative algorithm, consisting of the state estimation step followed by the parameter estimation step. At each iteration, a sequence of estimated states is used to calculate the parameter set and the resulting parameters is used to generate a new sequence of states for the next iteration. The procedure stops when the algorithm converges. The algorithm is initialized using a bounded random state sequence which guarantees the initial parameter set forms a stable kernel.
\label{StateAndParameterEstimationSection}

In this section we describe the procedure for estimating the states that scale the field basis functions, $\mathbf{x}_t$, the connectivity kernel parameters, $\boldsymbol \theta$, and the synaptic time constant, $\xi$. The estimation process consists of a two part iterative algorithm, consisting of the state estimation step followed by the parameter estimation step. At each iteration, the state estimate sequence is used to calculate the parameter set and the resulting parameters are used to generate a new state sequence for the next iteration. The procedure stops when the algorithm converges. The algorithm is initialized using a bounded random state sequence which guarantees the initial parameter set forms a stable kernel.

The state estimation technique uses the Unscented Rauch Tung Striebel Smoother (URTSS)~\cite{Sarkka2010}. The URTSS uses the Unscented Kalman Filter (UKF)~\cite{Julier1997, Merwe2003} in the forward iteration to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward iteration to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and the second order moments of the predicted state is captured by propagating the so-called sigma points through the state equation. The sigma points $\mathcal X_i$ are calculated using the unscented transform as follows

@@

where $(\sqrt{( L + \lambda)\mathbf P_x})_i$ is the $i$th column of the weighted matrix square root of  $\mathbf P_x$ and  $L$ is the dimension of the state space. The total number of sigma points is $2L+1$. The scaling parameter $\lambda$ is defined as 

@@

where $\alpha$, $\beta$ and $\kappa$ are constants (see ~\cite{Haykin2001} for more details). It should be noted that $\bar x$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$ and $\mathbf{P}_x$ represents the corresponding covariance matrix from the filtering or smoothing. The sigma vectors are propagated through the system equations and weighted to form the predicted mean and covariance. The weights are calculated by 

@@

Since the observation equation \ref{ObservationEquation} is linear the standard Kalman Filter update equations can be used to update the predicted states. The state estimates from the forward filtering are used to form a new set of sigma points for the smoother, as described above. In order to compute the smoother gain the predicted cross-covariance matrix of the states needs to be calculated. A summary of the URTSS procedure is given in algorithm~\ref{UKFAlgorithm}. It should be noted that as the disturbance, $e_t\left(\mathbf{r}\right)$ and the measurement noise $ \boldsymbol{\varepsilon}_t$ are additive the additive form of the URTS can be used , otherwise the state vector should be augmented with the noise terms. Using the additive form reduces the dimension and the number of sigma points used in the algorithm.

@@

% The accuracy of the approximation is linked to the time step used in the temporal discretisation of the system, where the state transition appears more linear with a finer resolution. 
Alternatives to the UKF are the Extended Kalman Filter (EKF)~\cite{Haykin2001} and the Sequential Monte Carlo (SMC) filter~\cite{doucet2001}. The EKF approximates the state transition equation by linearizing about the current state estimates (by calculating the Jacobian) at each time instance. The linearisation maintains the Gaussianity of the model. The unscented Kalman filter approximates the posterior state density by a Gaussian distribution using a minimal set of carefully chosen sigma points, whilst maintaining the nonlinearity in the system. This has been shown to give superior performance over the EKF in state estimation, as the EKF maintains a first order approximation where the UKF provides an approximation accurate at least to the second order. In addition, the computation of the Jacobian used in the EKF can be problematic. SMC filtering can theoretically provide an exact posterior state density for nonlinear systems. However, currently this method is not appropriate for our problem, due to computation demands.

%The method is appropriate for state estimation of stochastic nonlinear dynamical systems. 
\section{Results}\label{ResultsSection} The neural field model derived in section~\ref{NeuralModelSection} was used to generate the data for state and parameter estimation. All parameters for the model are given in Table~\ref{tab:Model Parameters}. An example of the simulated field can be seen in figure~\ref{fig:experimental design}~A. The Figure demonstrates the experimental design where the sensors were placed in the centre of the field (full width at half maximum). This design allowed for uncertainty around the edges of the sensor array, which would be typical from physical data, and alleviated problems associated with boundary conditions. Figure~\ref{fig:experimental design}~B. shows example data from the sensors of the model, and figure~\ref{fig:experimental design}~C. shows the power spectral density of the simulated time-series. The PSD shows the typical $1/f$ characteristics of intracranial EEG.

>>>>

The dynamic field must be spatially oversampled using a larger number of sensors then imposed by frequency analysis considerations, to use frequency analysis to determine the minimum distance and width of the basis functions. Following this, the distance between basis functions is usually larger then the distance between sensors, although this is not a requirement for using this estimation framework. For a spatially homogeneous isotropic fields, the basis functions can be placed on a regular grid. Following this, the knowledge of the distance between basis functions directly implies the total number of basis functions required.

% 
% Due to the high dimensionality of the brain and the current electrode systems, we can not expect to have more sensors than basis functions.
\section{State and Parameter Estimation}\label{StateAndParameterEstimationSection} In this section we describe the procedure for estimating the states, $\mathbf{x}_t$, that scale the field basis functions and the connectivity kernel parameters, $\boldsymbol \theta$, and the synaptic time constant, $\xi$. The estimation process is a two part iterative algorithm, consisting of the state estimation step followed by the parameter estimation step. At each iteration, the sequence of estimated states is used to update the parameter set, then resulting parameters are used to estimate a new state vector for the next iteration. The procedure stops when the parameters converge. The algorithm is initialised using a bounded random state sequence which guarantees the initial estimated parameter set forms a stable kernel.

The Unscented Rauch Tung Striebel Smoother (URTSS)~\cite{Sarkka2010} was used for the state estimation. The URTSS uses the unscented Kalman Filter (UKF)~\cite{Julier1997, Merwe2003} in a forward iteration to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward iteration to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and the second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. The sigma points, $\mathcal X_i$, are calculated using the unscented transform as follows

@@

where $\bar x$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$, $\mathbf{P}_x$ is the corresponding covariance matrix from the filtering or smoothing, $(\sqrt{( L + \lambda)\mathbf P_x})_i$ is the $i$th column of the weighted matrix square root of $\mathbf P_x$ and $L$ is the dimension of the state space. The total number of sigma points is $2L+1$. The scaling parameter $\lambda$ is defined as 

@@

where $\alpha$, $\beta$ and $\kappa$ are constants (see ~\cite{Haykin2001} for more details). The sigma vectors are propagated through the system equations and weighted to form the predicted mean and covariance. The weights are calculated by 

@@

Since the observation equation is linear (equation~\ref{ObservationEquation}) the standard Kalman Filter update equations are  used to correct the predicted states. The state estimates from the forward filtering are used to form a new set of sigma points for the smoother, as described above. To compute the smoother gain, the predicted cross-covariance matrix of the states is also calculated. A summary of the URTSS procedure is given in Algorithm~\ref{UKFAlgorithm}. It should be noted that as the disturbance, $e_t\left(\mathbf{r}\right)$ and the measurement noise $ \boldsymbol{\varepsilon}_t$ are additive the additive form of the URTS can be used , otherwise the state vector should be augmented with the noise terms. Using the additive form reduces the dimension and the number of sigma points used in the algorithm.

@@

% The accuracy of the approximation is linked to the time step used in the temporal discretisation of the system, where the state transition appears more linear with a finer resolution. 
Alternatives to the UKF are the Extended Kalman Filter (EKF)~\cite{Haykin2001} and the Sequential Monte Carlo (SMC) filter~\cite{doucet2001}. The EKF approximates the state transition equation by linearising about the current state estimates (by calculating the Jacobian) at each time instance. The linearisation maintains the Gaussianality of the model. The unscented Kalman filter approximates the posterior state density by a Gaussian distribution using a minimal set of carefully chosen sigma points, whilst maintaining the nonlinearity in the system. This has been shown to give superior performance over the EKF in state estimation, as the EKF maintains a first order approximation where the UKF provides an approximation accurate at least to the second order. In addition, the computation of the Jacobian used in the EKF can be problematic. SMC filtering can theoretically provide an exact posterior state density for nonlinear systems. However, currently this method is not appropriate for our problem, due to computation demands.

%The method is appropriate for state estimation of stochastic nonlinear dynamical systems. 
\section{Results}\label{ResultsSection} The neural field model derived in Section~\ref{NeuralModelSection} was used to generate the data for state and parameter estimation. All parameters for the model are given in Table~\ref{tab:Model Parameters}. An example of the simulated field can be seen in figure~\ref{fig:experimental design}~A. The Figure demonstrates the experimental design where the sensors were placed in the centre of the field (full width at half maximum). This design allowed for uncertainty around the edges of the sensor array, which would be typical from physical data, and alleviated problems associated with boundary conditions. Figure~\ref{fig:experimental design}~B. shows example data from the sensors of the model, and figure~\ref{fig:experimental design}~C. shows the power spectral density of the simulated time-series. The PSD shows the typical $1/f$ characteristics of intracranial EEG.
